Hippocampal replay of experience at real-world speeds

### 2020-10-19 Article.tex
Update on Overleaf.

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow10-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay. One might therefore conclude that the retrieval of stored representation during SWRs is limited to sequences of locations, and that all such retrieval events are sequential in nature. Our subjective experience is that we can retrieve memories of locations without having to mentally traverse a long distance. If the rodent hippocampus is capable of that sort of retrieval, then one might expect to see SWRs where the neural activity corresponds not to a very rapid trajectory through space but instead to either a stable pattern associated with a single location or perhaps a pattern more similar to one seen during a real experience. Importantly, this sort of activity could still be useful, as it would take place during the short duration of an SWR (\texttildelow100 ms) and could thus support rapid retrieval and use of representations.

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time, which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulate 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct a 280 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of movement dynamics that could be expressed during SWRs, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

We find that spatially coherent events indeed have lower HPD sizes than events with fragmented dynamics. Figure \ref{4}A and D show two example SWRs that are classified as having stationary and continuous dynamics, respectively. The most probable positions (as estimated by the model) at each time step in these SWRs is concentrated in a small portion of the track and correspondingly, the HPD size is small throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the most probable positions are more spatially diffuse and the HPD size is much larger. The HPD size also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}B, where the HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examine the average HPD size for each SWR, grouped by dynamic, we find a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the average HPD size for each SWR was much smaller than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). Additionally, the HPD size for the unclassified SWRs was similarly large. We note here that while the size of the HPD will be influenced by the dynamic estimated from the data, it remains possible for continuous or stationary dynamics to correspond to a large HPD region (see outliers in Figure \ref{4}E), indicating less certainty for those specific events. In general, if the data does not exhibit strong place specific firing, the HPD size will be large, regardless of dynamic classification. To show this, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes have much larger HPD size than the real data (epoch \#1: 55 cm vs. 229 cm, p=0.02, real vs. shuffled; epoch \#2: 72 cm vs. 231 cm, p=0.02, real vs. shuffled; one-sided Mann-Whitney-U, \ref{fig:Figure4-Figure supplement 1}).

>>>>

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow10-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay \cite{MichonPostlearningHippocampalReplay2019, KaeferReplayBehavioralSequences2020, ShinDynamicsAwakeHippocampalPrefrontal2019, DavidsonHippocampalReplayExtended2009}. One might therefore conclude that the retrieval of stored representation during SWRs is limited to sequences of locations, and that all such retrieval events are sequential in nature. Our subjective experience is that we can retrieve memories of locations without having to mentally traverse a long distance. If the rodent hippocampus is capable of that sort of retrieval, then one might expect to see SWRs where the neural activity corresponds not to a very rapid trajectory through space but instead to either a stable pattern associated with a single location or perhaps a pattern more similar to one seen during a real experience. Importantly, this sort of activity could still be useful, as it would take place during the short duration of an SWR (\texttildelow100 ms) and could thus support rapid retrieval and use of representations.

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time, which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the state space model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulate 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct a 280 ms spiking sequence (Figure \ref{1}A) and apply our model to the sequence. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of movement dynamics that could be expressed during SWRs, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

We find that spatially coherent events indeed have lower HPD sizes than events with fragmented dynamics. Figure \ref{4}A and D show two example SWRs that are classified as having stationary and continuous dynamics, respectively. The most probable positions (as estimated by the model) at each time step in these SWRs is concentrated in a small portion of the track and correspondingly, the HPD size is small throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the most probable positions are more spatially diffuse and the HPD size is much larger. The HPD size also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}B, where the HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examine the average HPD size for each SWR, grouped by dynamic, we find a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the average HPD size for each SWR was much smaller than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). Additionally, the HPD size for the unclassified SWRs was similarly large. We note here that while the size of the HPD will be influenced by the dynamic estimated from the data, it remains possible for continuous or stationary dynamics to correspond to a large HPD region (see outliers in Figure \ref{4}E), indicating less spatial certainty for those specific events. In general, if the data does not exhibit strong place specific firing, the HPD size will be large, regardless of dynamic classification. To show this, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes have much larger HPD size than the real data (epoch \#1: 55 cm vs. 229 cm, p=0.02, real vs. shuffled; epoch \#2: 72 cm vs. 231 cm, p=0.02, real vs. shuffled; one-sided Mann-Whitney-U, \ref{fig:Figure4-Figure supplement 1}).
